Generalized Rybicki Press algorithm

### 2014-09-27 paper/ESS.tex
Final version

\title{Generalized Rybicki Press scheme}
\subtitle{$\mathcal{O}(N)$ direct solver and determinant computation of general semi-separable matrices}

@@

\begin{abstract}
This article discusses a more general and numerically stable Rybicki Press scheme, which enables inverting and computing determinants of covariance matrices, whose elements are sums of exponentials. The algorithm is true in exact arithmetic and relies on introducing new variables and corresponding equations, thereby converting the matrix into a banded matrix of larger size. Linear complexity banded algorithms for solving linear systems and computing determinants on the larger matrix enable linear complexity algorithms for the initial semi-separable matrix as well. Benchmarks provided illustrate the linear scaling of the algorithm.
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{$\mathcal{O}(N)$ direct solver and determinant computation of general semi-separable matrices}{Generalized Rybicki Press scheme}

@@

\begin{itemize}
\item
A new $\mathcal{O}(N)$ direct solver for semi-separable matrices is obtained by embedding the semi-separable matrix into a larger banded matrix.
\item
The determinant of these semi-separable matrix is shown to equal to the determinant of the larger banded matrix, thereby enabling computing determinants of these semi-separable matrices at a computational cost of $\mathcal{O}(N)$. This is the first scheme for computing the determinants for a general semi-separable matrix.
\item
A numerically stable generalized Rybicki Press scheme is derived using these ideas. To be specific, fast direct solver and stable, fast determinant computation (both scaling as $\mathcal{O}(n)$) for covariance matrix of the form:
\begin{align}
A_{ij} = \sum_{l=1}^p \alpha_l \exp\left(-\beta_l \vert t_i - t_j \vert\right)
\label{eqn_covariance_matrix}
\end{align}
where $i,j \in \{1,2,\ldots,n\}$, the points $t_i$ are distinct and are distributed on an interval. The covariance matrix in Equation~\eqref{eqn_covariance_matrix} is frequently encountered in computational statistics in the context of Continuous time AutoRegressive-Moving-Average (abbreviated as CARMA) models~\cite{brockwell2002introduction, brockwell2001levy, brockwell1994continuous}.
\item
Another advantage of this scheme from a practical view-point is that the scheme relies only on sparse linear algebra and thereby can easily use the existing mature sparse linear algebra libraries.
\end{itemize}

@@

\textbf{Acknowledgements}: The author would like to thank Christopher S. Kochanek for initiating the conversation on generalized Rybicki Press scheme and David W. Hogg for putting in touch with Christopher S. Kochanek. The author was supported in part by the Applied Mathematical Sciences Program of the U.S. Department of Energy under Contract DEFGO288ER25053, Office of the Assistant Secretary of Defense for Research and Engineering and AFOSR under NSSEFF Program Award FA9550-10-1-0180.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.5]{./images/1termsemiseparable/1termsemiseparable.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix obtained from a rank $1$ semi-separable matrix where $N=8$. The color code is as follows: }
\label{fig_rank1_semiseparable}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%												%
%		SECTION: GENERAL SEMI-SEPARABLE			%
%												%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sparse embedding of a general semi-separable matrix}
Let $A$ be a $N \times N$ matrix, whose semi-separable rank is $p$, i.e., we have
\begin{align}
A(i,j) & =
\begin{cases}
a_{ii} & \text{ if }i=j\\
\displaystyle\sum_{l=1}^p u_j^{(l)} v_i^{(l)} & \text{ if }i>j\\
\displaystyle\sum_{l=1}^p u_i^{(l)} v_j^{(l)} & \text{ if }i<j\\
\end{cases}
\end{align}
where $i,j \in \{1,2,\ldots,N\}$. We then add the following variables $r_2^{(p)},r_3^{(p)},\ldots,r_N^{(p)}$ and $l_1^{(p)},l_2^{(p)},\ldots,l_{N-1}^{(p)}$ as before. However, unlike earlier and not surprisingly, these new variables $r_i^{(p)}$'s and $l_j^{(p)}$'s will be vectors of length $p$. Let $U_k^{(p)} = \begin{bmatrix} u_k^{(1)} & u_k^{(2)} & u_k^{(3)} & \cdots & u_k^{(p)}\end{bmatrix}$ and $V_k^{(p)} = \begin{bmatrix} v_k^{(1)} & v_k^{(2)} & v_k^{(3)} & \cdots & v_k^{(p)}\end{bmatrix}$. We then have the following relations for the additional vector variables.
\begin{align}
r_N^{(p)} = V_N^Tx_N\\
l_1^{(p)} = U_1^Tx_1
\end{align} and
\begin{align}
r_k^{(p)} & = V_k^T x_k + r_{k+1}^{(p)}\\
l_k^{(p)} & = U_k^T x_k + l_{k-1}^{(p)}
\end{align}
where $k \in \{2,\ldots,N-1\}$. Hence, we now have $(2p+1)N-2p$ variables (this includes the $N$ $x_k$'s, $N-1$ vector variables $r_k^{(p)}$ and $l_k^{(p)}$ of length $p$) and $(2p+1)N-2p$ equations relating them. Therefore, we end up with a $((2p+1)N-2p) \times ((2p+1)N-2p)$ extended sparse matrix, whose bandwidth is $(2p+1)$. This is illustrated in Figure~\ref{fig_rankm_semiseparable} for $10 \times 10$ semi-separable matrix, whose semi-separable rank is $4$.
\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.225]{./images/mtermsemiseparable/mtermsemiseparable_matrix.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$ and $p=4$. The color code is as follows: }
\label{fig_rankm_semiseparable}
\end{figure}

@@

The extended system, denoted by $A_{ex}$ on appropriate reordering of rows and columns can be written as
\begin{align}
P_1A_{ex} P_2
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix} & =
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix}
\end{align}
where $P_1$, $P_2$ are permutation matrices, the matrix $I_1$ is a highly sparse lower-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the lower-triangular part (the precise location is important for determinant computations as we will see later), the matrix $I_2$ is a highly sparse upper-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the upper-triangular part and $D$ is a diagonal matrix. The first set of rows, i.e., $\begin{bmatrix} I_1 & 0 & U_a\end{bmatrix}$, correspond to adding the variables $l_k^{(p)}$, i.e., $l_k^{(p)} = U_k^Tx_k + l_{k-1}^{(p)}$. The next set of rows, i.e., $\begin{bmatrix} 0 & I_2 & V_a\end{bmatrix}$, correspond to adding the variables $r_k^{(p)}$, i.e., $r_k^{(p)} = V_k^Tx_k + r_{k+1}^{(p)}$. The last set of rows, i.e., $\begin{bmatrix} V_b & U_b & D\end{bmatrix}$, correspond to the initial set of equations with the $l_k^{(p)}$'s and $r_k^{(p)}$'s introduced.
We then have
\begin{align}
\det(P_1A_{ex}P_2) & =
\underbrace{
\det\left(
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\right) =
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right)
\det\left(D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}
\right)
}_{\text{Block determinant formula}}
\end{align}
Now note that $\det(I_1) = 1 = \det(I_2)$, due to the fact that $I_1$ and $I_2$ are triangular matrices with $1$'s on the diagonal. Hence,
\begin{align}
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right) & = \det(I_1) \det(I_2) = 1 \times 1 = 1
\end{align}
Further, note that the matrix $D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}$ is the Schur complement obtained by eliminating the variables $l_i^{(p)}$, $r_i^{(p)}$ and hence is the initial dense matrix $A$ we began with. Hence, we have
\begin{align}
\det(A_{ex}) & = \pm \det(A)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%										%
%		SECTION: RYBICKI PRESS			%
%										%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reinterpretation of Rybicki Press scheme in terms of sparse embedding}

We will first naively reinterpret the Rybicki Press scheme in terms of the extended sparse matrix algebra. Recall that the Rybicki Press scheme inverts a correlation matrix $A$ given by Equation~\eqref{eqn_Rybicki_Press_matrix}.
\begin{align}
A(i,j) & = \exp\left(-\beta \lvert t_i - t_j \rvert\right)
\label{eqn_Rybicki_Press_matrix}
\end{align}
where $t_i$'s lies on an interval and are monotone. The original Rybicki Press algorithm relies on the fact that the inverse of $A$ happens to be a tridiagonal matrix. The key ingredient of their algorithm is the following property of exponentials:
\begin{align}
\exp\left(\beta (t_i-t_j) \right)\exp\left(\beta (t_j-t_k) \right) = \exp\left(\beta (t_i-t_k)\right)
\end{align}
In our sparse interpretation as well, we will use this property to recognize that the matrix $A$ is a semi-separable matrix, whose semi-separable rank is $1$. This can be seen by setting $u_k = \exp(\beta t_k)$ and $v_k = \exp(-\beta t_k)$. This then gives us ($i<j$) that $A(i,j) = u_i v_j = \exp(\beta t_i) \exp(-\beta t_j) = \exp(\beta(t_i-t_j))$ and similarly for $i>j$. This shows that the matrix $A$ is semi-separable with semi-separable rank $1$. Hence, we can mimic the same approach as in the earlier sections to obtain an $\mathcal{O}(N)$ algorithm. However, there is an issue that needs to be addressed from a numerical perspective. If the $t_i$'s are spread over a large interval, then $u_i$ is exponentially large, while $v_i$ is exponentially small, and hence embedding into a sparse matrix as such could prove to be a catastrophic leading to underflow and overflow of the relevant entries. This issue though can be circumvented by a suitable analytic preconditioning, by an appropriate change of variables. This is illustrated for a $4 \times 4$ linear system. We will use the notation $t_{ij}$ to denote $\lvert t_i - t_j\lvert$. The linear equation is
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & \exp(-\beta t_{13}) & \exp(-\beta t_{14})\\
\exp(-\beta t_{12}) & 1 & \exp(-\beta t_{23}) & \exp(-\beta t_{24})\\
\exp(-\beta t_{13}) & \exp(-\beta t_{23}) & 1 & \exp(-\beta t_{34})\\
\exp(-\beta t_{14}) & \exp(-\beta t_{24}) & \exp(-\beta t_{34}) & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\x_2\\x_3\\x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\b_2\\b_3\\b_4
\end{bmatrix}
\end{align}
Now lets introduce the additional variables as follows:
\begin{align}
r_4 & = x_4\\
r_3 & = x_3 + \exp(-\beta t_{34})r_4\\
r_2 & = x_2 + \exp(-\beta t_{23})r_3
\end{align}
\begin{align}
l_2 & = x_1\exp(-\beta t_{12})\\
l_3 & = (x_2+l_2)\exp(-\beta t_{23})\\
l_4 & = (x_3+l_3)\exp(-\beta t_{34})
\end{align}
The equations then become
\begin{align}
x_1 + \exp(-\beta t_{12}) r_2 & = b_1\\
l_2 + x_2 + \exp(-\beta t_{23}) r_3 & = b_2\\
l_3 + x_3 + \exp(-\beta t_{24}) r_4 & = b_3\\
l_4 + x_4 & = b_4
\end{align}
Embedding this in an extended sparse matrix, we obtain
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
\exp(-\beta t_{12}) & 0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & -1 & 0 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & \exp(-\beta t_{23}) & \exp(-\beta t_{23}) & 0 & -1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & -1 & 0 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 1 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & \exp(-\beta t_{34}) & \exp(-\beta t_{34}) & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & -1 & 0 & 1\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\ r_2\\ l_1\\ x_2\\ r_3\\ l_2\\ x_3\\ r_4\\ l_3\\ x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\ 0\\ 0\\ b_2\\ 0\\ 0\\ b_3\\ 0\\ 0\\ b_4
\end{bmatrix}
\end{align}

@@

%%%%%%%%%%%%%%%%%
%	FIGURE 1	%
%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
\subfigure[Assembly time versus system size]{
\includegraphics{./images/results/assembly_versus_N.pdf}
}\subfigure[Factor time versus system size]{
\includegraphics{./images/results/factor_versus_N.pdf}
}
\subfigure[Solve time versus system size]{
\includegraphics{./images/results/solve_versus_N.pdf}
}
\subfigure[Error versus system size]{
\includegraphics{./images/results/error_versus_N.pdf}
}
\caption{Scaling of the algorithm with system size. From the benchmarks, it is clear that the computational cost for the fast algorithm scales as $\mathcal{O}(N)$ for assembly, factorization and solve stages, where $N$ is the number of unknowns. The maximum residual is less than $10^{-13}$ even for a system with million unknowns.}
\label{figure_benchmark1}
\end{figure}

@@

%%%%%%%%%%%%%%%%%
%	FIGURE 2	%
%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
\subfigure[Assembly time versus number of exponentials]{
\includegraphics{./images/results/assembly_versus_m.pdf}
}\subfigure[Factor time versus number of exponentials]{
\includegraphics{./images/results/factor_versus_m.pdf}
}
\subfigure[Solve time versus number of exponentials]{
\includegraphics{./images/results/solve_versus_m.pdf}
}
\subfigure[Error versus number of exponentials]{
\includegraphics{./images/results/error_versus_m.pdf}
}
\caption{Scaling of the fast algorithm with the number of exponentials added. From the benchmarks, it is clear that the computational cost for the fast algorithm scales as $\mathcal{O}(p)$ for assembly and solve stages, while it scales as $\mathcal{O}(p^2)$ for the factorization stage, where $p$ is the number of exponentials (equivalently the semi-separable rank). The maximum residual is less than $10^{-13}$ almost always.}
\end{figure}

>>>>

\title{Generalized Rybicki Press algorithm}
\subtitle{$\mathcal{O}(N)$ direct solver and determinant computation of general semi-separable matrices}

@@

\begin{abstract}
This article discusses a more general and numerically stable Rybicki Press algorithm, which enables inverting and computing determinants of covariance matrices, whose elements are sums of exponentials. The algorithm is true in exact arithmetic and relies on introducing new variables and corresponding equations, thereby converting the matrix into a banded matrix of larger size. Linear complexity banded algorithms for solving linear systems and computing determinants on the larger matrix enable linear complexity algorithms for the initial semi-separable matrix as well. Benchmarks provided illustrate the linear scaling of the algorithm.
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{$\mathcal{O}(N)$ direct solver and determinant computation of general semi-separable matrices}{Generalized Rybicki Press algorithm}

@@

\begin{itemize}
\item
A new $\mathcal{O}(N)$ direct solver for semi-separable matrices is obtained by embedding the semi-separable matrix into a larger banded matrix.
\item
The determinant of these semi-separable matrix is shown to equal to the determinant of the larger banded matrix, thereby enabling computing determinants of these semi-separable matrices at a computational cost of $\mathcal{O}(N)$. This is the first algorithm for computing the determinants for a general semi-separable matrix.
\item
A numerically stable generalized Rybicki Press algorithm is derived using these ideas. To be specific, fast direct solver and stable, fast determinant computation (both scaling as $\mathcal{O}(n)$) for covariance matrix of the form:
\begin{align}
A_{ij} = \sum_{l=1}^p \alpha_l \exp\left(-\beta_l \vert t_i - t_j \vert\right)
\label{eqn_covariance_matrix}
\end{align}
where $i,j \in \{1,2,\ldots,n\}$, the points $t_i$ are distinct and are distributed on an interval. The covariance matrix in Equation~\eqref{eqn_covariance_matrix} is frequently encountered in computational statistics in the context of Continuous time AutoRegressive-Moving-Average (abbreviated as CARMA) models~\cite{brockwell2002introduction, brockwell2001levy, brockwell1994continuous}.
\item
Another advantage of this algorithm from a practical view-point is that the algorithm relies only on sparse linear algebra and thereby can easily use the existing mature sparse linear algebra libraries.
\end{itemize}

@@

\textbf{Acknowledgements}: The author would like to thank Christopher S. Kochanek for initiating the conversation on generalized Rybicki Press algorithm and David W. Hogg for putting in touch with Christopher S. Kochanek. The author was supported in part by the Applied Mathematical Sciences Program of the U.S. Department of Energy under Contract DEFGO288ER25053, Office of the Assistant Secretary of Defense for Research and Engineering and AFOSR under NSSEFF Program Award FA9550-10-1-0180.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.5]{./images/1termsemiseparable.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix obtained from a rank $1$ semi-separable matrix where $N=8$. The color code is as follows: }
\label{fig_rank1_semiseparable}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%												%
%		SECTION: GENERAL SEMI-SEPARABLE			%
%												%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Sparse embedding of a general semi-separable matrix}
Let $A$ be a $N \times N$ matrix, whose semi-separable rank is $p$, i.e., we have
\begin{align}
A(i,j) & =
\begin{cases}
a_{ii} & \text{ if }i=j\\
\displaystyle\sum_{l=1}^p u_j^{(l)} v_i^{(l)} & \text{ if }i>j\\
\displaystyle\sum_{l=1}^p u_i^{(l)} v_j^{(l)} & \text{ if }i<j\\
\end{cases}
\end{align}
where $i,j \in \{1,2,\ldots,N\}$. We then add the following variables $r_2^{(p)},r_3^{(p)},\ldots,r_N^{(p)}$ and $l_1^{(p)},l_2^{(p)},\ldots,l_{N-1}^{(p)}$ as before. However, unlike earlier and not surprisingly, these new variables $r_i^{(p)}$'s and $l_j^{(p)}$'s will be vectors of length $p$. Let $U_k^{(p)} = \begin{bmatrix} u_k^{(1)} & u_k^{(2)} & u_k^{(3)} & \cdots & u_k^{(p)}\end{bmatrix}$ and $V_k^{(p)} = \begin{bmatrix} v_k^{(1)} & v_k^{(2)} & v_k^{(3)} & \cdots & v_k^{(p)}\end{bmatrix}$. We then have the following relations for the additional vector variables.
\begin{align}
r_N^{(p)} = V_N^Tx_N\\
l_1^{(p)} = U_1^Tx_1
\end{align} and
\begin{align}
r_k^{(p)} & = V_k^T x_k + r_{k+1}^{(p)}\\
l_k^{(p)} & = U_k^T x_k + l_{k-1}^{(p)}
\end{align}
where $k \in \{2,\ldots,N-1\}$. Hence, we now have $(2p+1)N-2p$ variables (this includes the $N$ $x_k$'s, $N-1$ vector variables $r_k^{(p)}$ and $l_k^{(p)}$ of length $p$) and $(2p+1)N-2p$ equations relating them. Therefore, we end up with a $((2p+1)N-2p) \times ((2p+1)N-2p)$ extended sparse matrix, whose bandwidth is $(2p+1)$. This is illustrated in Figure~\ref{fig_rankm_semiseparable} for $10 \times 10$ semi-separable matrix, whose semi-separable rank is $4$.
\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.225]{./images/mtermsemiseparable_matrix.pdf}
\end{center}
\begin{center}
\includegraphics[scale=1]{./images/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$ and $p=4$. The color code is as follows: }
\label{fig_rankm_semiseparable}
\end{figure}

@@

The extended system, denoted by $A_{ex}$ on appropriate reordering of rows and columns can be written as
\begin{align}
P_1A_{ex} P_2
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix} & =
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\begin{bmatrix}
l_1^{(p)}\\
l_2^{(p)}\\
\vdots\\
l_{N-1}^{(p)}\\
r_2^{(p)}\\
r_3^{(p)}\\
\vdots\\
r_N^{(p)}\\
x_1\\
x_2\\
\vdots\\
x_N
\end{bmatrix}
\end{align}
where $P_1$, $P_2$ are permutation matrices, the matrix $I_1$ is a highly sparse lower-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the lower-triangular part (the precise location is important for determinant computations as we will see later), the matrix $I_2$ is a highly sparse upper-triangular matrix with $1$'s on the diagonal and $-1$'s at a few places in the upper-triangular part and $D$ is a diagonal matrix. The first set of rows, i.e., $\begin{bmatrix} I_1 & 0 & U_a\end{bmatrix}$, correspond to adding the variables $l_k^{(p)}$, i.e., $l_k^{(p)} = U_k^Tx_k + l_{k-1}^{(p)}$. The next set of rows, i.e., $\begin{bmatrix} 0 & I_2 & V_a\end{bmatrix}$, correspond to adding the variables $r_k^{(p)}$, i.e., $r_k^{(p)} = V_k^Tx_k + r_{k+1}^{(p)}$. The last set of rows, i.e., $\begin{bmatrix} V_b & U_b & D\end{bmatrix}$, correspond to the initial set of equations with the $l_k^{(p)}$'s and $r_k^{(p)}$'s introduced.
We then have
\begin{align}
\det(P_1A_{ex}P_2) & =
\underbrace{
\det\left(
\begin{bmatrix}
I_1 & 0 & U_a\\
0 & I_2 & V_a\\
V_b & U_b & D
\end{bmatrix}
\right) =
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right)
\det\left(D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}
\right)
}_{\text{Block determinant formula}}
\end{align}
Now note that $\det(I_1) = 1 = \det(I_2)$, due to the fact that $I_1$ and $I_2$ are triangular matrices with $1$'s on the diagonal. Hence,
\begin{align}
\det\left(
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}
\right) & = \det(I_1) \det(I_2) = 1 \times 1 = 1
\end{align}
Further, note that the matrix $D-
\begin{bmatrix}
U_a\\
V_a
\end{bmatrix}
\begin{bmatrix}
I_1 & 0\\
0 & I_2
\end{bmatrix}^{-1}
\begin{bmatrix}
U_b & V_b
\end{bmatrix}$ is the Schur complement obtained by eliminating the variables $l_i^{(p)}$, $r_i^{(p)}$ and hence is the initial dense matrix $A$ we began with. Hence, we have
\begin{align}
\det(A_{ex}) & = \pm \det(A)
\end{align}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%										%
%		SECTION: RYBICKI PRESS			%
%										%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Reinterpretation of Rybicki Press algorithm in terms of sparse embedding}

We will first naively reinterpret the Rybicki Press algorithm in terms of the extended sparse matrix algebra. Recall that the Rybicki Press algorithm inverts a correlation matrix $A$ given by Equation~\eqref{eqn_Rybicki_Press_matrix}.
\begin{align}
A(i,j) & = \exp\left(-\beta \lvert t_i - t_j \rvert\right)
\label{eqn_Rybicki_Press_matrix}
\end{align}
where $t_i$'s lies on an interval and are monotone. The original Rybicki Press algorithm relies on the fact that the inverse of $A$ happens to be a tridiagonal matrix. The key ingredient of their algorithm is the following property of exponentials:
\begin{align}
\exp\left(\beta (t_i-t_j) \right)\exp\left(\beta (t_j-t_k) \right) = \exp\left(\beta (t_i-t_k)\right)
\end{align}
In our sparse interpretation as well, we will use this property to recognize that the matrix $A$ is a semi-separable matrix, whose semi-separable rank is $1$. This can be seen by setting $u_k = \exp(\beta t_k)$ and $v_k = \exp(-\beta t_k)$. This then gives us ($i<j$) that $A(i,j) = u_i v_j = \exp(\beta t_i) \exp(-\beta t_j) = \exp(\beta(t_i-t_j))$ and similarly for $i>j$. This shows that the matrix $A$ is semi-separable with semi-separable rank $1$. Hence, we can mimic the same approach as in the earlier sections to obtain an $\mathcal{O}(N)$ algorithm. However, there is an issue that needs to be addressed from a numerical perspective. If the $t_i$'s are spread over a large interval, then $u_i$ is exponentially large, while $v_i$ is exponentially small, and hence embedding into a sparse matrix as such could prove to be a catastrophic leading to underflow and overflow of the relevant entries. This issue though can be circumvented by a suitable analytic preconditioning, by an appropriate change of variables. This is illustrated for a $4 \times 4$ linear system. We will use the notation $t_{ij}$ to denote $\lvert t_i - t_j\lvert$. The linear equation is
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & \exp(-\beta t_{13}) & \exp(-\beta t_{14})\\
\exp(-\beta t_{12}) & 1 & \exp(-\beta t_{23}) & \exp(-\beta t_{24})\\
\exp(-\beta t_{13}) & \exp(-\beta t_{23}) & 1 & \exp(-\beta t_{34})\\
\exp(-\beta t_{14}) & \exp(-\beta t_{24}) & \exp(-\beta t_{34}) & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\x_2\\x_3\\x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\b_2\\b_3\\b_4
\end{bmatrix}
\end{align}
Now lets introduce the additional variables as follows:
\begin{align}
r_4 & = x_4\\
r_3 & = x_3 + \exp(-\beta t_{34})r_4\\
r_2 & = x_2 + \exp(-\beta t_{23})r_3
\end{align}
\begin{align}
l_2 & = x_1\exp(-\beta t_{12})\\
l_3 & = (x_2+l_2)\exp(-\beta t_{23})\\
l_4 & = (x_3+l_3)\exp(-\beta t_{34})
\end{align}
The equations then become
\begin{align}
x_1 + \exp(-\beta t_{12}) r_2 & = b_1\\
l_2 + x_2 + \exp(-\beta t_{23}) r_3 & = b_2\\
l_3 + x_3 + \exp(-\beta t_{24}) r_4 & = b_3\\
l_4 + x_4 & = b_4
\end{align}
Embedding this in an extended sparse matrix, we obtain
\begin{align}
\begin{bmatrix}
1 & \exp(-\beta t_{12}) & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
\exp(-\beta t_{12}) & 0 & -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\
0 & -1 & 0 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & 1 & 1 & \exp(-\beta t_{23}) & 0 & 0 & 0 & 0 & 0\\
0 & 0 & \exp(-\beta t_{23}) & \exp(-\beta t_{23}) & 0 & -1 & 0 & 0 & 0 & 0\\
0 & 0 & 0 & 0 & -1 & 0 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & 1 & 1 & \exp(-\beta t_{34}) & 0 & 0\\
0 & 0 & 0 & 0 & 0 & \exp(-\beta t_{34}) & \exp(-\beta t_{34}) & 0 & -1 & 0\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & -1 & 0 & 1\\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 1\\
\end{bmatrix}
\begin{bmatrix}
x_1\\ r_2\\ l_1\\ x_2\\ r_3\\ l_2\\ x_3\\ r_4\\ l_3\\ x_4
\end{bmatrix}
=
\begin{bmatrix}
b_1\\ 0\\ 0\\ b_2\\ 0\\ 0\\ b_3\\ 0\\ 0\\ b_4
\end{bmatrix}
\end{align}

@@

%%%%%%%%%%%%%%%%%
%	FIGURE 1	%
%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
\subfigure[Assembly time versus system size]{
\includegraphics{./images/assembly_versus_N.pdf}
}\subfigure[Factor time versus system size]{
\includegraphics{./images/factor_versus_N.pdf}
}
\subfigure[Solve time versus system size]{
\includegraphics{./images/solve_versus_N.pdf}
}
\subfigure[Error versus system size]{
\includegraphics{./images/error_versus_N.pdf}
}
\caption{Scaling of the algorithm with system size. From the benchmarks, it is clear that the computational cost for the fast algorithm scales as $\mathcal{O}(N)$ for assembly, factorization and solve stages, where $N$ is the number of unknowns. The maximum residual is less than $10^{-13}$ even for a system with million unknowns.}
\label{figure_benchmark1}
\end{figure}

@@

%%%%%%%%%%%%%%%%%
%	FIGURE 2	%
%%%%%%%%%%%%%%%%%
\begin{figure}[!htbp]
\subfigure[Assembly time versus number of exponentials]{
\includegraphics{./images/assembly_versus_m.pdf}
}\subfigure[Factor time versus number of exponentials]{
\includegraphics{./images/factor_versus_m.pdf}
}
\subfigure[Solve time versus number of exponentials]{
\includegraphics{./images/solve_versus_m.pdf}
}
\subfigure[Error versus number of exponentials]{
\includegraphics{./images/error_versus_m.pdf}
}
\caption{Scaling of the fast algorithm with the number of exponentials added. From the benchmarks, it is clear that the computational cost for the fast algorithm scales as $\mathcal{O}(p)$ for assembly and solve stages, while it scales as $\mathcal{O}(p^2)$ for the factorization stage, where $p$ is the number of exponentials (equivalently the semi-separable rank). The maximum residual is less than $10^{-13}$ almost always.}
\end{figure}
